Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-20 inference.tex
Sec. 5

({\sf exists} translates to $\exists$ in first-order logic, as
formalized in Fig.~\ref{fig:logic}), which is enough to prove that
$\C{a.bal + pc*a.bal}\ge 0$, and verify that \C{add\_interest}
preserves the positive balance invariant.

@@

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similar.

The first-order encoding of a bind expression describes the semantics
of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain predicates
$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
under a new binder that binds $x$, free variable sequence of $s_2$ is
$\nubar,x$. For technical reasons, we substitute a fresh $a$ for $x$,
making the sequence $\nubar,a$. The set $s$ is encoded as a a new
uninterpreted predicate $g$ indexed by $s$'s free variables
($\nubar$). Since the set denoted by $g$ is the result of the bind
$s_1 \bind \lambda x.s_2$, a first-order constraint defining the bind
operation (as described above) is generated. The constraint relates the
predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to
the uninterpreted predicate $g$ that represents $s$.

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic)
is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in \S\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.

>>>>

({\sf exists} has the same semantics as $\exists$, as formalized in
Fig.~\ref{fig:logic}), which is enough to prove that $\C{a.bal +
pc*a.bal}\ge 0$, and verify that \C{add\_interest} preserves the
positive balance invariant.

@@

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similarly.

The first-order encoding of a bind expression describes the semantics
of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain $(\phi_1,\G_1)$
and $(\phi_2,\G_2)$, respectively. Since $s_2$ is under a new binder
that binds $x$, free variable sequence of $s_2$ is $\nubar,x$. In the
interest of hygiene, we substitute a fresh $a$ for $x$, making the
sequence $\nubar,a$. The set $s$ is encoded as a a new uninterpreted
predicate $g$ indexed by $s$'s free variables ($\nubar$). Since the
set denoted by $g$ is the result of the bind $s_1 \bind \lambda
x.s_2$, a first-order constraint defining the bind operation (as
described above) is generated. The constraint relates the predicates
$\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to the
uninterpreted predicate $g$ that represents $s$. The constraint is
assigned a name ($\pi$) so that it can be enabled only for the choice
of free variables ($\nubar$) for which $g$ needs to be defined;
observe that the lambda returned by $\mssemof{\nubar}{s_1 \bind
\lambda x.s_2}$ asserts $\pi(\vbar)$ only in conjunction with
$g(\vbar,r)$, for some $r$.

The first-order encoding of the $\existsl(\stg,\phi,s)$ expression
eliminates the existential by Skolemizing it. Skolemizing is the
process of substituting an existentially bound $x$ in
$\phi_x\in\Prop^1$ with $f(\nubar)$, where $f$ is a fresh
uninterpreted function, and $\nubar$ are the free variables in
$\phi_x$ bound by enclosing universal quantifiers. Due to the
decidability restrictions (Sec.~\ref{sec:decidability}), the only
uninterpreted functions we admit in our logic are boolean (i.e.,
predicates/relations). Consequently, Skolemizing an existentially
bound $\stg$ with $f(\nubar)$ is not really an option in our case.
Instead, we define $f$ as an uninterpreted relation that behaves as a
partial function:
\begin{smathpar}
  \begin{array}{c}
    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
    \Rightarrow a = b
  \end{array}
\end{smathpar}
We then replace the existentially bound $\stg$ with a new universally
bound $a$ in $\phi$ and $s$, and define the set $[a/\stg]s$ under the
condition that $f(\nubar,a) \conj [a/\stg]\phi$ holds. Since the
existence of a $\stg$ that satisfies $\phi$ is axiomatized tp
define $s$, if the aforementioned condition does not hold, i.e., if
there does exist such a $\stg$, we let the context derive $\bot$.

\noindent {\bf Example} Let us reconsider the  TPC-C \C{new\_order}
transaction from Sec.~\ref{sec:motivation}. Recall that the state
transformer ($\T$) for the foreach loop shown in
Fig.~\ref{fig:foreach_code} is (\C{k1}, \C{k2}, and \C{k3} are
constants):
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Where:
\begin{smathpar}
  \begin{array}{lcl}
    \F_U & = & \lambda(\stl,\stg).~ \stg \bind(\lambda s. 
                \itel{{\sf table}(s) = \C{Stock} \conj 
                  s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
         &   & \hspace*{0.9in}}
                     {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                                 \C{s\_d\_id}=s.\C{s\_d\_id};\,
                                 \C{s\_qty} = \C{k1}\rangle \}\\
         &   & \hspace*{0.9in}} {\emptyset}) \\
    \F_I & = & \lambda(\stl,\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,
                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}\\
  \end{array}
\end{smathpar}
For any $\stl$ and $\stg$, $\F_U(\stl,\stg)$ and $\F_I(\stl,\stg)$ are
expressions in $\SL$, so can be translated to FOL by the encoding
algorithm in Fig.~\ref{fig:logic}. Since the iteration variable
\C{item\_req} occurs free in these expressions, the appropriate
application of the encoding algorithm is
$\mssemof{\C{item\_req}}{\F_U(\stl,\stg)}$ and
$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which result in
$(\phi_U,\G_U)$ and $(\phi_I,\G_I)$, respectively, where $\phi_U$,
$\phi_I$, $\G_U$, $\G_I$ are as shown below:
\begin{smathpar}
  \begin{array}{lcl}
    \phi_U & = & \forall \C{item\_req}.\forall s. \forall s'.~  \\
    & & \hspace*{0.1in}\pi(\C{item\_req}) \Leftrightarrow(s \in \stg) \conj 
        (\itec{{\sf table}(s)=\C{Stock} \wedge 
              s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
    & & \hspace*{1.45in}}
              {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                            \C{s\_d\_id}=s.\C{s\_d\_id};\,
                            \C{s\_qty} = \C{k1}\rangle\\
    & & \hspace*{1.45in}} {\bot}) \\
    & & \hspace*{1.2in} \Leftrightarrow g(\C{item\_req},s')\\
    \G_U & = & \lambda (\C{item\_req},r).~ \pi(\C{item\_req}) 
        \wedge g(\C{item\_req},r)\\
    \phi_I & = & \top \\ 
    \G_U & = & \lambda (\C{item\_req},r).~ 
        r = \langle \C{ol\_o\_id}=\C{k2},\,
                    \C{ol\_d\_id}=\C{k3},\, 
                    \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                    \C{ol\_qty}=\C{item\_req.ol\_qty}
             \rangle\\
  \end{array}
\end{smathpar}

\subsection{Decidability}
\label{sec:decidability} %this label is referenced earlier.

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item All quantifiers are universal, and appear only at the prenex
    position, i.e., at the beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic) is
known to be decidable~\cite{z3epr}. The language of encoding, however,
is a combination of {\sf EPR} with (a). $\Prop^0$, the theory from
which quantifier-free propositions ($\varphi$) that encode object
language expressions are drawn, and (b).  $\Prop^1$, the theory from
which invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in \S\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.
